Reconstructing the last known movements of one of Nature's giants

### 2019-02-14 manuscript/PeerJ/blue-new.tex
Edits from CT

The spatial ecology of rare, migratory oceanic mammals such as blue
whales (\textit{Balaenoptera musculus}) is difficult to study directly. 
Recent advances in satellite telemetry have improved our understanding of individual-level migratory behavior, but tags generally last only a few months, and cannot provide retrospective information. 
Incrementally-grown tissues provide an alternative means of reconstructing individual-level movements over long timescales, but inferring spatio-temporal information from biochemical tracers is challenging. 
Here we present a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to infer individual-level animal movements. 
We use our new method to reconstruct multiple years of movement histories of individual modern and historic blue whales, including an iconic blue whale which stranded in 1891 and is now on display at the Natural History Museum, London.
Our method unlocks the information recorded in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.

@@

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \citep{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \citep{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time \citep{hobson2008tracking}. 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline) \citep{west2006stable,mcmahon2015millennial}. 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues.

Here we introduce a new approach to infer location from the isotopic compositions of sequentially-sampled incrementally-grown tissues by drawing on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Fig. S3), coupled to an agent-based model of whale movements (see Supplementary Methods; Table S1; Fig. S5).
We use our coupled models to simulate time series of carbon stable isotope compositions expected under different movement behaviours. 
We then compare simulated isotopic profiles with the measured profiles to identify potential movement histories most consistent with the measured isotopic profile.

We apply our method to blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \citep{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass \citep{goldbogen2015}.

Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes \citep{huckgaete2018}. 
As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions \citep{branch2007}. 
Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or winter
areas) \citep{busquets2017estimating}. 
The northeast Pacific population of blue whales is perhaps the best studied in terms of individual-level movements with a recent history of deployment of over 177 satellite tags \citep{irvine2017quantifying, mate2007evolution}.
In the northeast Pacific, satellite tags show individuals occupying most of the exclusive economic zone (EEZ) and into the Gulf of Alaska \citep{irvine2017quantifying} with greatest overlapping areas in the southern half of the EEZ, along the continental slope and shelf region between the California Channel Islands and Cape Mendocino. 
Winter feeding is less well characterized, but winter feeding grounds are known
in the Gulf of California and in the eastern tropical Pacific, specifically the Costa Rica Dome where production is maintained by seasonal upwelling.
However, most satellite deployments on blue whales report movement data for periods of time under six months \citep{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas. 
The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \citep{irvine2017quantifying} and in the North Atlantic is 177 days \citep{lesage2017foraging}.

@@

Here we infer individual level high-resolution movement records for seven blue whales for which baleen stable isotope data were available.
Isotope data for six whales stranding on the west coast of the USA or Mexico were reported in \cite{busquets2017estimating}, and we report new isotopic data from a whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM).
The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
Baleen from the NHM whale therefore yields information about migratory behaviors in North Atlantic blue whales around the peak of industrial whaling.

@@

\subsection{Stable isotope extractions from baleen}\label{stable-isotope-extractions-from-baleen}

@@

\subsection{Time calibrating stable isotope
profiles}\label{time-calibrating-stable-isotope-profiles}

Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}. 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Supplementary Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.

\subsection{Baseline isotope
comparisons}\label{baseline-isotope-comparisons}

Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize
the isotopic composition of phytoplankton expected in different
potential foraging grounds (Fig. S3). 
Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Fig. S3). \(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \citep{magozzi2017using,yool2013medusa}. 
Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig. S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{magozzi2017using,laws1995dependence}. 
Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
The model simulations used are forced with decadal climatological data from 2000-2010 \citep{magozzi2017using}. 
We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values caused by release of fossil carbon into the atmosphere (Suess effect) does not influence inferences about location or movement. 
Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters, therefore we are confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data.

We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date. 
Average values were weighted according to the proportional plankton biomass estimated for each month. 
Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
The degree of such trophic fractionation is unclear, however and we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate. 
We assume opportunistic ingestion of protein throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.

\subsection{Agent-based whale movement
model}\label{agent-based-whale-movement-model}

@@

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, or foraging).
Northerly migrations were possible only in spring, and southerly migrations in autumn. 
Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
(ii) Sea surface temperature \citep{yool2013medusa} ($^{\circ}$C). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.  
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\citep{bathy}, extracted via the marmap package in R \citep{marmap}). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \citep{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

For Pacific whales where satellite data offers a rich view of individual movements, we coded our movement models to reflect either residency in southern waters (between Baja California and the Costa Rica Dome) or seasonal latitudinal movements within the California current system. 
We simulated 250 movement tracks of four years duration. 
We allowed start and end dates to vary by aligning measured profiles to sections of simulated profiles of identical lengths with sequentially changing start dates and saved best fitting models.

For the historic North Atlantic whale, we simulated the isotopicexpression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Supplemental Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.
Measured tracks can only be replicated by combination of residency and seasonal migration with latitudinal migrations limited to the last four
years of simulations. 
For a full description of the model and its parameters see Supplementary Methods: Fig. S5 and Table S1. 
We simulated seven years of whale movements 1000 times and two years of residency 30 times for each residency hotspot (Fig. S4).
We simulated whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations.

At each time point and location in the simulations, we extracted time averaged \(\delta^{13}\)C values from the spatio-temporal model \citep{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas (Figs X, \ref{fig2}, and S4).
We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile of the blue whale (Figure \ref{fig1}).

R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \citep{github})

\section{Results and Discussion}

\subsection{Matching model to Pacific data.}

Estimates of baleen growth rates for the six modern whales from the northeast Pacific were inferred from cyclical fluctuations in nitrogen isotope ratios and reported in \cite{busquets2017estimating}. 
We used these growth rates to reconstruct a potential timeline for baleen isotope records.

% Pacific whales figure
\begin{figure}
 \centering
  \includegraphics[width = \linewidth]{figures/figure-Pacific-draft.eps}
  \caption{Correlations among simulated $\delta^{13}$C from the best fitting migratory movement models (grey lines, right hand y-axis) and $\delta^{13}$C from baleen (black line, left hand y-axis) for six Pacific blue whales taken from \cite{busquets2017estimating}. 
  Simulated $\delta^{13}$C values are six month moving average values for the time series of simulated plankton $\delta^{13}$C values in that location, reflecting temporal integration of phytoplankton $\delta^{13}$C values within the food chain before ingestion by the whale as krill. 
  The end points of the simulations and empirical data have been aligned to coincide.
}
  \label{pacific}
\end{figure}

In four of the six cases, measured baleen carbon isotope profiles could be replicated from model simulations with relatively high precision (Fig. \ref{pacific}
; best fit r\textsuperscript{2} values ranging from 0.43 to 0.74).
Whales D and F showed similar carbon isotope records and were best fit by the same movement simulation. 
We extracted monthly locations associated with the best fitting simulation models for each whale (Fig. \ref{pacific}).
Best fitting simulation models for whales D and F imply a relatively southerly distribution with few locations north of Baja, California and relatively limited seasonal latitudinal migration. 
The measured carbon isotope record for whale A was best fitted by simulations with a southerly distribution, but with an additional strong seasonal latitudinal migration, and with more northerly locations between April and October. 
Whale C was best fit by models with relatively northerly locations and strong seasonal latitudinal migration. 
We were unable to find convincing fits to baleen profiles for whales B and E suggesting that longer simulations with a greater variation in movements were needed. 
The inferred movement records are consistent with the known spatial distributions and relatively high level of individual variation in spatial distributions among blue whales in the Pacific northwest \citep{irvine2017quantifying, mate2007evolution}.

\subsection{Atlantic whale baleen measured isotopic record}

For Hope, the historic North Atlantic whale, \(\delta^{15}\)N values record regular cyclical fluctuations throughout the length of the baleen plate, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods). 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm\(y^{- 1}\), we reconstructed a timeline for \(\delta^{13}\)C and \(\delta^{15}\)N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891). 
The \(\delta^{13}\)C and \(\delta^{15}\)N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated \(\delta^{13}\)C values, and relatively low \(\delta^{15}\)N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1890) \(\delta^{15}\)N values are relatively elevated and \(\delta^{13}\)C values are relatively depleted with coincident cyclical fluctuations in both \(\delta^{13}\)C and \(\delta^{15}\)N values. 
In the last year of life the cyclical pattern is disrupted, with constant low \(\delta^{13}\)C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive \(\delta^{13}\)C values in the second half of 1890. 
The final three months of the record show a progressive fall in \(\delta^{13}\)C values (Fig. \ref{fig1}; Figs. S1 and S2).

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Supplemental Figures S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

>>>>

The spatial ecology of rare, migratory oceanic mammals such as blue
whales (\textit{Balaenoptera musculus}) is difficult to study directly. 
Recent advances in satellite telemetry have improved our understanding of individual-level migratory behavior, but tags generally last only a few months, and cannot provide retrospective information. 
Incrementally-grown tissues provide an alternative means of reconstructing individual-level movements over long timescales, but inferring spatio-temporal information from biochemical tracers is challenging. 
Here we outline a new approach combining stable isotope analyses of incrementally-grown tissues with simulation models to test hypotheses relating to individual-level animal movements. 
We illustrate our new method by identifying the most likely multi-year movement history of a historic blue whale that stranded in 1891 and is now on display at the Natural History Museum, London. 
Our method unlocks movement information coded in biochemical compositions of incremental tissues including those housed in historic collections, and provides inferences that are complimentary to and comparable with tag-derived movement data.

@@

An alternative technique for investigating animal movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \citep{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \citep{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
A large literature describes the use of stable isotope markers to link the composition of a tissue to a geographic source at a single point in time \citep{hobson2008tracking}. 
Relating the isotopic compositions of marine animal tissues to the likely location of tissue growth is complicated by a lack of knowledge of spatial variations in the isotopic composition of diet (the isotopic baseline) \citep{west2006stable,mcmahon2015millennial}. 
This uncertainty is compounded when multiple samples are taken through time in the same individual, as temporal variation in the isotopic baseline must also be considered explicitly. 
Consequently, relatively little attention has been given to the potential to infer individual movement histories by reconstructing time series of multiple discrete origins from sequentially sampled incrementally-grown tissues 
% [trueman st john glew, Tokyo eel, darnaude, brennan?].

Recently, mechanistic models have been developed providing relatively accurate predictions of isotopic variability at high temporal and spatial resolution. 
These models can be coupled to Lagrangian or agent-based models of animal movement to predict isotopic compositions or trajectories expected to be recorded in animal tissues under differing movement scenarios. 
Here we illustrate the potential of this conceptual approach by assessing the relative likelihood of differing, plausible hypotheses of movement behavior for a single individual blue whale based on carbon isotope compositions sampled across a baleen plate.  
We draw on newly developed models predicting spatio-temporal variation in phytoplankton stable carbon isotope composition at global and monthly resolution \citep{magozzi2017using} (Figure S3), coupled to an agent-based model of whale movements (see Table S1; Figure S5). 
We specifically ask whether the isotopic profile observed in baleen can be simulated using agent based movement models informed by current understanding of animal movements and isotopic variability.

We apply our method a blue whales (\textit{Balaenoptera musculus}), the largest animal to have ever lived. 
Mysticete whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \citep{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior typically reflecting multiple years of life of an individual whale. 
Baleen is worn away at the tips over time, so a baleen plate reflects the most recent years of life, and rarely records an individual's entire lifespan. 
Among the mysticete whales, blue whales are a particularly attractive target for model-based isotope movement work as they have a consistent, low trophic level diet and feed continuously through the year, likely driven by high energetic costs of maintaining extreme muscle mass \citep{goldbogen2015}.

Similar to other large balaenopterid whales, blue whales are generally assumed to conduct annual migrations between high and low latitudes \citep{huckgaete2018}. 
As blue whales feed throughout the year to accommodate energetic costs, migration routes and feeding areas for blue whales are thought to be shaped by the year-round location of highly productive regions \citep{branch2007}. 
Recent studies tracking movements of individual blue whales primarily in the northeast Pacific have demonstrated high levels of among-individual variation in movement histories including suspending migration for opportunistic feeding and skipped migration (i.e. year round residency in either summer or winter
areas) \citep{busquets2017estimating}. 
However, most satellite deployments on blue whales report movement data for periods of time under six months \citep{heide2001new, silva2013north, bailey2009behavioural, lesage2017foraging, irvine2017quantifying}, and are unable to show individual movements between summer and winter feeding or breeding areas. 
The longest period of continuous monitoring via satellite tagging of an individual blue whale that we are aware of in the northeast Pacific is 504 days \citep{irvine2017quantifying} and in the North Atlantic is 177 days \citep{lesage2017foraging}.

@@

Here we aim to show that stable carbon isotope tracers can be used to test hypotheses about individual-level movement behavior, and can be applied to recent or historic archived samples. 
We apply our method to infer the most likely movement history for a blue whale stranding off the coast of Wexford Ireland in March 1891, and currently on display in the Natural History Museum, London (NHM). 
The NHM whale ``Hope'' is a female estimated to be a young adult at least 15 years at death. 
Baleen from the NHM whale therefore yields information about about a single blue whale living in the North Atlantic around the peak of industrial whaling.

@@

\subsection{Stable isotope extractions from baleen}
\label{stable-isotope-extractions-from-baleen}

@@

\subsection{Time calibrating stable isotope profiles}
\label{time-calibrating-stable-isotope-profiles}

Seasonal variations in isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates \citep{hobson1998stable,busquets2017estimating}. 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the youngest 70 - 18cm of the plate (behavioural phase two). 
We calculated isotopic periodicity within the NHM baleen sample using Fourier Transform analysis \citep{cardona2017temporal} (Figure S1), revealing a consistent growth rate of 13.5$cmy^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cmy^{-1}$ estimated by \cite{busquets2017estimating}.  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891.

\subsection{Baseline isotope comparisons}
\label{baseline-isotope-comparisons}

Isotope-enabled biogeochemical ocean models \citep{magozzi2017using,schmittner2016complementary} were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds (Figure S3). 
Annual average \(\delta^{15}\)N POM (particulate organic matter) values were provided by C.J. Somes (\textit{pers.comm}) based on a 5\({}^{\circ}\) resolution biogeochemical model (Figure S3). 
\(\delta^{13}\)C POM values were simulated at 1\({}^{\circ}\) and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model \citep{magozzi2017using,yool2013medusa}. 
Simulated \(\delta^{15}\)N POM values are relatively positive in the northeast Atlantic north of c. 60\({}^{\circ}\)N, and relatively negative in the central and southern North Atlantic. 
Annual average \(\delta^{13}\)C POM values largely vary with latitude, with more negative values in more northerly regions. 
In the central North Atlantic, \(\delta^{13}\)C POM values are relatively positive in the west, reflecting warm gulf stream waters (Figure S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature \citep{magozzi2017using,laws1995dependence}. 
Thus temporal variations in \(\delta^{13}\)C POM values are superimposed on latitudinal gradients. 
The scale and nature of temporal variation in \(\delta^{13}\)C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in \(\delta^{13}\)C POM values linked to strongly seasonal phytoplankton growth dynamics.
The model simulations used are forced with decadal climatological data from 2000-2010 \citep{magozzi2017using}.

The isotopic composition of phytoplankton is influenced by the release of fossil carbon with low \(\delta^{13}\)C values into the atmosphere (Suess effect) and the effect of increased concentrations of dissolved CO2 on isotopic fractionation during phytoplankton growth 
%(Young et al 2013)
, potentially complicating the application of a model trained on recent data to historic samples. 
Ocean basin scale spatio-temporal differences in \(\delta^{13}\)C values largely reflect relative differences in sea surface temperatures between trophic, temperate and arctic waters and over seasons, and we assume that any regional variation in isotopic effects due to CO\textsubscript{2}-influenced changes in phytoplankton composition and growth rate are minor compared to latitudinal and seasonal variations in \(\delta^{13}\)C POM values. 
We do not draw inferences about location from absolute \(\delta^{13}\)C values and thus the reduction in oceanic \(\delta^{13}\)C POM values does not directly influence inferences about location or movement.  
We are therefore confident that the \cite{magozzi2017using} model can be applied to interpret historic baleen isotope data at least at ocean-basin scale spatial resolution.

We used \(\delta^{13}\)C POM values modeled at monthly resolution to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours. 
The stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth. 
Assimilation of carbon into krill tissues will dampen the temporal variability seen in POM, effectively producing a temporal average over the timescale of isotopic turnover within krill. 
We estimate turnover to be complete between two and four months and therefore we resampled the \(\delta^{13}\)C POM values in each one degree cell to reflect an average of isotopic compositions in phytoplankton in the two months prior to the sampling date. 
Average values were weighted according to the proportional plankton biomass estimated for each month. 
Carbon isotope values are also likely to be fractionated during transfer from plankton to krill as \textsuperscript{12}C is preferentially lost through respiration. 
The degree of such trophic fractionation is unclear, however and as we do not draw interpretations based on absolute \(\delta^{13}\)C values, rather on the relative \(\delta^{13}\)C values across the length of the baleen plate, we do not need to quantify this trophic enrichment effect. 
We assume that carbon available for synthesis of keratin is drawn from carbon released by respiration of diet captured opportunistically throughout the year \citep{baines2017autumn,silva2013north,visser2011timing,busquets2017estimating, lesage2017foraging, bailey2009behavioural, branch2007, huckgaete2018}.

\subsection{Agent-based whale movement model}
\label{agent-based-whale-movement-model}

We simulated the likely isotopic expression associated with different movement behaviours by building an agent-based movement model within the R coding environment, where movement behavior is influenced by sea surface temperature and phytoplankton biomass estimates provided by NEMO-MEDUSA \citep{yool2013medusa}, and bathymetry from the General Bathymetric Chart of the Oceans (GEBCO) \citep{bathy}, extracted via the marmap package in R \citep{marmap} (Figure S5).
% check figure S5

@@

In the models, the likelihood of moving, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and linear distance of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, or foraging, fixed according to month as defined by the operator).
Directly, northerly migrations were coded to occur in spring, and southerly migrations in autumn.
Foraging was possible at any time of year, and was triggered when whales encountered high concentrations of plankton.
(ii) Sea surface temperature \citep{yool2013medusa} ($^{\circ}$C). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters.  
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset\citep{bathy}, extracted via the marmap package in R \citep{marmap}). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration ($mmolNm^{-3}$, for combined diatom and non-diatom communities \citep{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state. 
At each daily step, the probability of movement, movement direction and distance traveled are sampled from probability distributions to allow individual variation.

Initial boundary conditions are defined with a maximum temperature of 25$^{\circ}$C and minimum temperature of 3$^{\circ}$C. 
The likelihood of movement (i.e. whether to move or not from the current location) is sampled from a binomial distribution with the probability of movement influenced by behavioural state and external conditions. 
The maximum daily movement distance permitted in each behavioural state is defined as a random sample of a Gaussian distribution with specified mean and standard deviation (see Table S1).

The agent-based mode provides daily location data points which are then used to extract time averaged \(\delta^{13}\)C values from the isotopic extension to the NEMO-MEDUSA model and assembled to build simulated isotopic profiles across baleen plates.
We created experimental movement models by manipulating the relative duration of the foraging and migratory behavioural states to produce simulated records of isotopic composition of carbon in baleen expected under differing movement trajectories. 
We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, Canaries/Azores and Mid Atlantic Ridge, and the Cape Verde/Mauritanian upwelling area \citep{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Figure S4); (b) seasonal migration between high sub-Arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes.

We simulated two years of residency 30 times for each residency hotspot (Figure S4). 
We simulated seven years of whale movements 1200 times, then excluded simulations where the virtual whale stranded before reaching the 3019 days of the baleen record, leaving 1049 simulations. 
We then compared the simulated stable isotope profiles (Figure \ref{fig3}) to the profile measured in the blue whale baleen (Figure \ref{fig1}) with simple linear regressions.

For a full description of the model and its parameters see Figure S5 and Table S1. 
R code for all analyses is available from GitHub (https://github.com/nhcooper123/blue-whale-bes; \citep{github})

\section{Results}

\subsection{Baleen stable isotope profiles}
The baleen plate yielded 97 discrete samples of baleen.
\(\delta^{15}\)N values measured in the baleen plate display regular cyclical fluctuations throughout the length of the plate. Fourier analysis (periodogram function in the R package TSA \citep{Chan:2012aa}) revealed a strong periodic repetition with a 13.3cm periodicity, with a mean spacing of 13.5cm, assumed to represent annual periodicity. 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm y$^{-1}$, we reconstructed a timeline for \(\delta^{13}\)C and \(\delta^{15}\)N fluctuations in the baleen over seven full years of the whale’s life (early 1884 - spring 1891).

\(\delta^{13}\)C values are relatively high and constant in the oldest (most distal) 35cm of the baleen plate, associated with a weakening of the periodic fluctuation seen in \(\delta^{15}\)N values, and an overall reduction in \(\delta^{15}\)N values. 
This is followed by a clear change towards repeated fluctuations in \(\delta^{13}\)C values in the middle 50cm of the baleen plate, with a similar periodicity of 13.5cm (Figure \ref{fig1}). 
A second, distinct change in the pattern of \(\delta^{13}\)C values along the baleen plate occurs at around 84cm from the distal end, with a transition to uniform, relatively low \(\delta^{13}\)C values followed by an abrupt transition to positive \(\delta^{13}\)C values and a decline in \(\delta^{13}\)C values in the most recent (most gingival) 3cm of the record.

The \(\delta^{13}\)C and \(\delta^{15}\)N profiles therefore divide the isotopic record into two distinct phases that we assume reflect changes in the whale’s behaviour. 
In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated \(\delta^{13}\)C values, and relatively low \(\delta^{15}\)N values (Figure \ref{fig1}; Figures S1 and S2). 
In behavioural phase two (summer 1886 to spring 1890) \(\delta^{15}\)N values are relatively high and \(\delta^{13}\)C values are relatively low with coincident cyclical fluctuations in both \(\delta^{13}\)C and \(\delta^{15}\)N values.
In the last year of life the cyclical pattern is disrupted, with constant low \(\delta^{13}\)C values for approximately six months in the first half of 1890, before a rapid switch to relatively high \(\delta^{13}\)C values in the second half of 1890. 
The final three months of the record show a progressive fall in \(\delta^{13}\)C values (Figure \ref{fig1}; Figures S1 and S2). 
Cross-correlation analysis demonstrates a strong negative covariance between \(\delta^{13}\)C and \(\delta^{15}\)N values within behavioural phase two (Figure S2), but no relationship between \(\delta^{13}\)C and \(\delta^{15}\)N values exists during behavioural phase one.

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = 97 samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figures S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

\subsection{Agent-based whale movement model}
